---
title: "Class Project Report"
subtitle: "Tree Health: Penn State Altoona"
author: "Sarah Allen"
date: "December 8, 2024"
output: 
  html_notebook:
    fig_height: 3
    fig_width: 5
---

## Research / Guiding Questions

Which tree species on the Penn State Altoona campus has the highest incidence of causing damage due to its location and condition risk?
How do the at risk trees on the Penn State Altoona campus compare (e.g., species) to the tree species in parks in the city of Portland, Oregon? 
What else can R do to help us learn more about Altoona campus and Portland, Oregon trees?

```{r message=FALSE}
# clean up R environment
rm(list = ls())

# load necessary packages
library(tidyverse)
library(dplyr)
library(esquisse)
library(ggplot2)
library(tidyr)
library(stringr)
library(pdxTrees)  # Package contains secondary data source including information about trees in the city of Portland, Oregon, USA
```

```{r}
AltoonaTrees <- read.csv("Altoona_Trees_Assigned_Risk.csv") # Data that is NOT from an R package

head(AltoonaTrees)
str(AltoonaTrees)
```

## Become Acquainted with data source: Altoona Trees
Where: I have access as the Altoona representative for the Commonwealth Arboreta Network.
Who: Collected by Barlett Tree Experts.
When & Why: Collected in summer 2024 as part of a multi-campus tree inventory.
Cases: Each case represents an individual tree on the Altoona campus that poses some risk to people or property. 139 cases.  
Variables of interest:scientific name, DBH (diameter at breast height), condition, etc.

## Become Acquainted with data source: Portland Trees
Where: Data within an R package (pdxTrees). 
Who: Data collected by staff and volunteers. 
When & Why: Trees were inventoried in 2017, 2018, and 2019. 
Cases: Each case is one tree at a specific latitude and longitude found in a Portland, OR park. 25,534 cases. 
Variables of interest: scientific name, DBH (diameter at breast height), condition, etc.

## Altoona Trees: Summary statistics of DBH (diameter at breast height)

```{r}
# Using reduction functions to learn about DBH

# Obtain Mean DBH Value with Base R methods
mean_DBH_baseR <- mean(AltoonaTrees$DBH, na.rm = TRUE)
print(mean_DBH_baseR)

# Obtain Mean DBH Value Value with TidyVerse methods (creates a new dataframe)
mean_DBH <- 
  AltoonaTrees %>%
  summarize(mean_column = mean(DBH, na.rm = TRUE))

mean_DBH

# Determine min and max DBH values of Altoona Trees
min_DBH_baseR <- 
  min(AltoonaTrees$DBH, na.rm = TRUE)

min_DBH_baseR

max_DBH_baseR <- 
  max(AltoonaTrees$DBH, na.rm = TRUE)

max_DBH_baseR
```

## Altoona Trees: Plots

```{r fig.height=9, fig.width=9}

ggplot(AltoonaTrees) +
 aes(x = Condition.Class, y = DBH) +
 geom_col(fill = "#112446") +
 theme_minimal()

ggplot(AltoonaTrees) +
 aes(x = Overall.Tree.Risk.Rating, colour = Overall.Tree.Risk.Rating, group = Common.Name) +
 geom_bar(fill = "#112446") +
 scale_color_hue(direction = 1) +
 theme_minimal()

ggplot(AltoonaTrees) +
 aes(x = Family) +
 geom_bar(fill = "#112446") +
 theme_minimal()

# Filtering the AltoonaTrees dataset to focus on only the Fagaceae (oak and beech) plant family.
Altoona_Fagaceae <-   
  AltoonaTrees %>%
  filter(Family == "Fagaceae") 
  
ggplot(Altoona_Fagaceae) +
 aes(x = DBH, y = Primary.Target, fill = Scientific_Name) +
 geom_boxplot() +
 scale_fill_hue(direction = 1) +
 theme_classic() +
 theme(axis.text.y = element_text(size = 13L), axis.text.x = element_text(size = 13L), 
 legend.text = element_text(size = 13L), legend.title = element_text(size = 13L))
```
**Notes on Figures:** Trees that are in "Fair condition" on the Altoona campus have the largest Diameter at Breast Height (DBH). Most trees in the dataset have a moderate risk rating. There are very few trees with a high risk rating. The majority of the trees on the Penn State Atloona campus that are at risk of failure are in the family Fagaceae (oak and beech family). This makes sense as oak and beech trees are very abundant on the Penn State Altoona campus. The last figure focuses on just the family Fagaceae and what impact per DBH different species in that family may have. *Quercus alba* and *Quercus rubra* are the two most common species in the family on the Altoona campus. They have a wide range of sizes (represented by DBH) and are likely to cause damage to a variety of campus items if they fell.   

## View Second Data Source: Trees in parks in Portland, Oregon
From R package "pdxTrees"

```{r}
pdxTrees_parks <- get_pdxTrees_parks()
str(pdxTrees_parks)
head(pdxTrees_parks)
```
## Portland Trees: Summary statistics for Tree Height & DBH

```{r}
# Using reduction functions to learn about Portland Trees

mean_DBH_Portland <- 
  pdxTrees_parks %>%
  summarize(meanDBH = mean(DBH, na.rm = TRUE))

mean_DBH_Portland

mean_height_Portland <-
  pdxTrees_parks %>%
  summarize(meanHeight = mean(Tree_Height, na.rm = TRUE))

mean_height_Portland
```

## Informative Plots about Portland, Oregon park trees

```{r}
# esquisser(pdxTrees_parks)

ggplot(pdxTrees_parks) +
 aes(x = Condition, y = Tree_Height) +
 geom_col(fill = "#112446") +
 theme_minimal()

ggplot(pdxTrees_parks) +
 aes(x = Condition, y = DBH) +
 geom_col(fill = "#112446") +
 theme_minimal()

ggplot(pdxTrees_parks) +
 aes(x = Tree_Height, fill = Condition) +
 geom_histogram(bins = 30L) +
 scale_fill_hue(direction = 1) +
 theme_minimal() +
 facet_wrap(vars(Native))

# Filtering the Portland park dataset to focus only on the species Quercus rubra and Quercus alba that are also common on the Altoona campus.
Portland_Fagaceae <-   
  pdxTrees_parks %>%
  filter(Scientific_Name == "Quercus rubra" | Scientific_Name == "Quercus alba") 
  
ggplot(Portland_Fagaceae) +
 aes(x = DBH, y = Condition, fill = Scientific_Name) +
 geom_boxplot() +
 scale_fill_hue(direction = 1) +
 theme_classic() +
 theme(axis.text.y = element_text(size = 13L), axis.text.x = element_text(size = 13L), 
 legend.text = element_text(size = 13L), legend.title = element_text(size = 13L))
```
**Notes on Figures:** Trees of "Fair" condition are the most common in the Portland Parks. 
The third plot is comparing the number of trees of different heights and faceting by whether they are native or non-native. There are more non-native trees that are shorter, whereas native trees tend to be taller. The fourth plot indicates that there are some *Quercus rubra* that are larger than *Quercus alba* in Portland parks. In addition, the *Quercus alba* that are present are in fair condition, whereas the *Quercus rubra* trees span all condition classes.    

## Additional Information about data sources
Using more data verbs.

```{r}
AltoonaTreesSpecies <-
  AltoonaTrees %>%
  group_by(Scientific_Name) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

head(AltoonaTreesSpecies, n = 10) 

PortlandTreeSpecies <-
  pdxTrees_parks %>%
  group_by(Scientific_Name) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

head(PortlandTreeSpecies, n = 10)
```

**Notes on Comparing the Datasets:** Both datasets have multiple species of *Quercus* (oak) in the top 10. *Quercus rubra* is common to the top 10 of both datasets. Oaks are a very important hardwood in North America. The Altoona dataset does not have any softwoods in the top 10, but the most abudant species in Portland parks is a softwood *Pseudotsuga menziesii* (Douglas fir). Species of *Prunus* (i.e., cherries, plums) and *Acer* (maples) are also common in both locations.   

## Join Data Frames using the Scientific name of species

```{r}
BothLocations <-
  inner_join(AltoonaTrees, pdxTrees_parks, by = "Scientific_Name")

head(BothLocations)
str(BothLocations)
```
## Using Regular Expressions

```{r}
# Using Regular expressions with Portland trees dataset

Hybrid <-   # Creates a df that only includes hybrid species. 
  pdxTrees_parks %>%
  group_by(Scientific_Name) %>%
  filter(grepl("hybrid", Origin, ignore.case = TRUE))

glimpse(Hybrid)


```

## Wide vs. Narrow data

```{r}
# Using Spread or Gather to stack/unstack variables with Portland trees dataset


```

